Study of Intra- and Inter-Job Interference for Improving Allocation on HPC Systems

### 2015-10-13 submission.tex
chages based on MM's comments

It has been widely recognized that network contention between concurrently running jobs in HPC system is one of the primary causes of performance variability. Optimizing job allocation and avoiding network sharing are proven techniques to alleviating network contention and performance degradation. In this work, we propose a novel job communication pattern aware allocation strategy which serve as a critical module in our future HPC batch scheduler. The novelty of our new strategy is that it makes allocation decisions based on job’s communication pattern communication pattern in addition to system topology and resource availability information. The scheduler uses job communication patterns information to decide whether a job’s performance relies heavily on the network or not, and what the potential inter/intra-job interference would be for a given allocation. Thus, the scheduler can pick a preferable resource allocation based on a specific job's communication pattern. Using traces have been collected from three representative parallel applications, we validate the effectiveness of our new design. The idea presented in this paper is widely applicable: HPC parallel applications have common communication patterns, and allocating resource with communication pattern awareness can improve performance on systems with a variety of network topologies.

@@

The scale of supercomputers keeps growing to accommodate the the demand for computing power required by scientific research areas. Supercomputers contain hundreds of thousands of nodes, and serve as irreplaceable research vehicle for scientific problems with increasing size and complexity. Supercomputers are employed as shared resource to accommodate many jobs running concurrently\ref{fig: overview} and maintain high system utilization. These jobs share the system infrastructure such as network and I/O bandwidth, and inevitably there is contention over the shared resources. As supercomputers continue to evolve, these shared resources are increasingly the bottleneck for performance.

\begin{figure}[h!]
    \centering
    \begin{subfigure}[t]{0.2\textwidth}
        \centering
        \includegraphics[height=1.2in]{figs/goodallocation}
        \caption{Contiguous}
        \label{fig:overview_sub1}
    \end{subfigure}%
    \hspace{1em}%
    \begin{subfigure}[t]{0.2\textwidth}
        \centering
        \includegraphics[height=1.2in]{figs/badallocation}
        \caption{Non-contiguous}
        \label{fig:overview_sub2}
    \end{subfigure}%
   \caption{Indication of multiple jobs running concurrently with different allocations. Different job represented by specific color. a) shows the effect of contiguous allocation, which would alleviate the inter-job interference. b) shows non-contiguous allocation, which definitely introduce both intra and inter-job interference. }
   \label{fig: overview}
\end{figure}

@@

One of the most prominent  problems is network contention between concurrently running jobs. The network sharing between concurrently running jobs causes communication variability that results in jobs running slower\cite{abhinav-sc13}. The delay of currently running jobs increases the queueing time of the following submitted jobs, thus leading to low system throughput, low utilization\cite{jose-ipdps15}.

This adverse effect of network sharing can be mitigated by providing jobs with better node allocations. The batch scheduler is responsible for assigning job the amount of nodes it requires. Currently, two allocation strategies are commonly used on torus-connected systems. One is so called partition based allocation, where the scheduler assigns each user job a compact and contiguous set of computing nodes. IBM Blue Gene series systems use this strategy. This strategy favors application performance by providing application with isolated partitions and exclusive network connections, thus reducing network contention caused by concurrently running jobs. However, this strategy can cause internal fragmentation (when more nodes are allocated to a job than it requests) and external fragmentation (when sufficient nodes are available for a request, but they can not be allocated contiguously), therefore leading to low system utilization than otherwise possible. The other strategy is non-contiguous allocation, where free nodes are assigned to user jobs whether they are contiguous or not. Cray XT/XE series systems use this strategy. Non-contiguous allocation eliminates internal and external fragmentation as seen in partition-based systems, thereby leading to high system utilization. However, it introduces other problems, for example scattering application processes all over the system, causing both inter-job and intra-job contention. The non-contiguous node allocation can significantly reduce job performance, especially for those communication-intensive jobs.

@@

%=========come back and work from here=============
Rather than choose blindly between partition-based or non-contiguous allocation, the new allocation strategy choses resources based on a job's communication pattern. With detail analysis of application's communication patterns, a clear view of its communication topology is obtained shows which processes are tightly connected and conduct intensive communication. The ``local community" consists of such tightly coupled processes should get a compact allocation. Then, the allocation strategy only needs to guarantee the locality of the subset of the nodes where those ``local communities" would reside.

In this work, we focus on the study of impact of different allocation strategies on intra/inter-job interference on Torus network. Torus networks have been extensively used in the current generation of supercomputers because of their linear scaling on per-node cost and competitive communication performance. On the recent Top500 list, 6 of the top 10 supercomputers use a high-radix torus-interconnected network \cite{top500}. The current generation of IBM Blue Gene/Q (BG/Q) supercomputer, such as Mira at Argonne Nation Lab and Sequoia at  Lawrence Livermore National Laboratory, has its nodes connected in a 5D torus network \cite{bgq}. The K computer from Japan uses the “Tofu” system, which is a 6D mesh/torus topology \cite{tofu}. Titan, a Cray XK7 supercomputer located at the Oak Ridge Leadership Computing Facility (OLCF), has nodes connected in a 3D torus within the compute partition \cite{titan}. Although the analysis are based on torus network, the idea conveyed from this work is applicable to networks with different topologies.

We selected three signature applications from DOE Design Forward Project\cite{design forward webpage} as examples to conduct deep study about their communication patterns. We use a sophisticated simulation toolkit named CODES, Co-Design of Multi-layer Exascale Storage Architecture\cite{Jason-2011}, from Argonne National Laboratory as research vehicle to evaluate the performance of our communication pattern aware allocation policy on torus networks. We analysis each application's performance when it running on torus networks with different dimensionality and interconnected bandwidth. Then, we extend CODES with a Job Mapping API that support different allocation policies on torus network. Different allocation strategies on torus network are studied with the help of this new Job Mapping API. We analysis the intra/inter-job interference by simulating those three applications running exclusively or concurrently on torus network with different allocation strategies. According to results of our comprehensive exploration, making allocation with consideration about job's communication pattern is a very promising allocation strategy. We believe the idea presented in this work would be the guideline for the design of future HPC batch job scheduler and resource management module.

The rest of this paper organized as follows. Section \ref{sec:application study} gives a detailed study of the three applications from DOE Design Forward Project. Section \ref{sec:codes} talks about CODEs as research vehicle for our work. Section \ref{sec:torus} shows the performance analysis of three applications on different torus networks. Section \ref{sec:interference} provides detailed analysis about the intra/inter-job interference between three applications on torus network. Section \ref{sec:discussion} talks about the idea of smart allocation strategy with job communication pattern awareness. Section \ref{sec:related_work} talks about the related work in the area and conclusion will be presented in Section \ref{sec:conclusion}.

@@

\begin{figure}[t!]
    \centering
    \begin{subfigure}[t]{0.22\textwidth}
        \centering
        \includegraphics[height=1.2in]{figs/appstudy/amg/amg_pip}
        \caption{Communication Topology}
        \label{fig: amg communication topology}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{0.22\textwidth}
        \centering
        \includegraphics[height=1.2in]{figs/appstudy/amg/amg_data_transfer}
        \caption{Data Volume}
        \label{fig: amg data trans}
    \end{subfigure}
    \caption{AMG. (a) rank-to-rank communication topology graph. (b) the data amount send/received by each rank in AMG.}
\end{figure}

@@

\begin{figure}[t!]
    \centering
    \begin{subfigure}[t]{0.22\textwidth}
        \centering
        \includegraphics[height=1.2in]{figs/appstudy/cr/cr_pip}
        \caption{Communication Topology}
        \label{fig: cr communication topology}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{0.22\textwidth}
        \centering
        \includegraphics[height=1.2in]{figs/appstudy/cr/cr_data_transfer}
        \caption{Data Volume}
        \label{fig: cr data trans}
    \end{subfigure}
    \caption{CrystalRouter. (a) rank-to-rank communication topology graph. (b) the data amount send/received by each rank in CrystalRouter. }
\end{figure}

@@

\begin{figure}[t!]
    \centering
    \begin{subfigure}[t]{0.22\textwidth}
        \centering
        \includegraphics[height=1.2in]{figs/appstudy/mg/mg_pip}
        \caption{Communication Topology}
        \label{fig: mg communication topology}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{0.22\textwidth}
        \centering
        \includegraphics[height=1.2in]{figs/appstudy/mg/mg_data_transfer}
        \caption{Data Volume}
        \label{fig: mg data trans}
    \end{subfigure}
    \caption{MultiGrid. (a) rank-to-rank communication topology graph. (b) the data amount send/received by each rank in MultiGrid.}
\end{figure}

@@

\subsection{Configuration Study }
\label{sec: config study}

>>>>

Network contention between concurrently running jobs on HPC system is one of the primary causes of performance variability. Optimizing job allocation and avoiding network sharing are proven techniques to alleviate network contention and performance degradation. In this work, we propose a novel job communication pattern aware allocation strategy which serves as a critical module in our future HPC batch scheduler. The novelty of our new strategy is that it makes allocation decisions based on job’s communication pattern in addition to system topology and resource availability information. The scheduler uses job communication patterns information to decide whether its performance relies heavily on the network or not, and what the potential inter/intra-job interference would be for a given allocation. Thus, the scheduler can pick a preferable resource allocation based on a specific job's communication pattern. We validate the effectiveness of our new design using traces collected from three representative parallel applications. The idea presented in this paper is widely applicable: HPC parallel applications have common communication patterns, and allocating resource with communication pattern awareness can improve performance on these systems with a variety of network topologies.

@@

The scale of supercomputers keeps growing to accommodate the the demand for computing power required by scientific research areas. Supercomputers have hundreds of thousands of nodes, and serve as irreplaceable research vehicle for scientific problems with increasing size and complexity. Supercomputers are usually employed as a shared resource to accommodate many parallel applications (also known as jobs) running concurrently \cite{zhou-ipdps} and maintain high system utilization. These communication and I/O intensive jobs share the system infrastructure such as network and I/O bandwidth, and inevitably there is contention over these shared resources. As supercomputers continue to evolve, these shared resources are increasingly the bottleneck for performance.

\begin{figure}[h!]
    \centering
    \begin{subfigure}[t]{0.2\textwidth}
        \centering
        \includegraphics[height=1.2in]{figs/goodallocation}
        \caption{Contiguous}
        \label{fig:overview_sub1}
    \end{subfigure}%
    \hspace{1em}%
    \begin{subfigure}[t]{0.2\textwidth}
        \centering
        \includegraphics[height=1.2in]{figs/badallocation}
        \caption{Non-contiguous}
        \label{fig:overview_sub2}
    \end{subfigure}%
   \caption{Indication of multiple jobs running concurrently with different allocations. Each job is represented by a specific color. a) shows the effect of partition-based allocation, which reduce the inter-job interference. b) shows non-contiguous allocation, which introduces both intra and inter-job interferences. }
   \label{fig: overview}
\end{figure}

@@

One of the most prominent  problems is network contention among concurrently running jobs. The network sharing among concurrently running jobs causes communication variability that results in jobs running slower \cite{abhinav-sc13}. The delay of running jobs increases the queueing time of the following submitted jobs, thus leading to low system throughput and utilization \cite{jose-ipdps15}.

This adverse effect of network sharing can be mitigated by providing jobs with isolated allocations and exclusive network resource. The batch scheduler is responsible for assigning job the amount of nodes it requires. On the widely used torus-connected HPC systems \cite{bgq}\cite{tofu}\cite{titan}, two allocation strategies are commonly used. In the partition based allocation, used in Blue Gene series systems, he scheduler assigns each user job a compact and contiguous set of computing nodes. This strategy favors application performance by providing application with isolated partitions and exclusive network connections, thus reducing network contention caused by concurrently running jobs. However, this strategy can cause internal fragmentation (when more nodes are allocated to a job than it requests) and external fragmentation (when sufficient nodes are available for a request, but they can not be allocated contiguously), therefore leading to lower system utilization that is otherwise possible. The other strategy is non-contiguous allocation, used by Cray XT/XE series, in which free nodes are assigned to user jobs whether they are contiguous or not. Non-contiguous allocation eliminates internal and external fragmentation as seen in partition-based systems, thereby leading to high system utilization. However, it introduces other problems, for example scattering application processes all over the system, causing both inter-job and intra-job contention. The non-contiguous node allocation can significantly reduce job performance, especially for communication-intensive jobs.

@@

%=========come back and work from here=============
Rather than choosing blindly between partition-based or non-contiguous allocation, the new allocation strategy choses resources based on a job's communication pattern. With detail analysis of application's communication patterns, our proposed allocation strategy obtains a clear view as to which processes are tightly connected and conduct intensive communication. The ``neighborhood communication" consists of such tightly coupled processes should get a compact allocation. Then, the allocation strategy only needs to guarantee the locality of the subset of the nodes where those ``neighborhood communication" would reside.

In this work, we study the impact of different allocation strategies on intra/inter-job interference on Torus network. Torus networks have been extensively used in the current generation of supercomputers because of their linear scaling on per-node cost and competitive communication performance. On the recent Top500 list, 6 of the top 10 supercomputers use a high-radix torus-interconnected network \cite{top500}. The current generation of IBM Blue Gene/Q (BG/Q) supercomputer, such as Mira at Argonne Nation Laboratory and Sequoia at  Lawrence Livermore National Laboratory, has its nodes connected in a 5D torus network \cite{bgq}. The K computer from Japan uses the “Tofu” system, which is a 6D mesh/torus topology \cite{tofu}. Titan, a Cray XK7 supercomputer located at the Oak Ridge Leadership Computing Facility (OLCF), has nodes connected in a 3D torus within the compute partition \cite{titan}. Although the analysis are based on torus network, the idea conveyed from this work is applicable to networks with different topologies.

We selected three signature applications from DOE Design Forward Project\cite{design forward webpage} as examples to conduct detailed study about their communication patterns. We use a sophisticated simulation toolkit named CODES, Co-Design of Multi-layer Exascale Storage Architecture\cite{Jason-2011}, from Argonne National Laboratory as a research vehicle to evaluate the performance of our communication pattern aware allocation policy on torus networks. We analyze each application's performance running on torus networks with different dimensionality and interconnected bandwidth. Then, we extend CODES with a Job Mapping API that supports different allocation policies on torus network. Different allocation strategies on torus network are studied with the help of this new Job Mapping API. We analyze the intra/inter-job interference by simulating those three applications running exclusively or concurrently on torus network with different allocation strategies. According to results of our comprehensive exploration, making allocation with consideration about job's communication pattern is a very promising allocation strategy. We believe the idea presented in this work would be the guideline for the design of future HPC batch job scheduler and resource management module.

The rest of this paper organized as follows. Section \ref{sec:application study} gives a detailed study of the three applications from DOE Design Forward Project. Section \ref{sec:codes} talks about CODEs as research vehicle for our work. Section \ref{sec:config study} shows the performance analysis of three applications on different torus networks. Section \ref{sec:interference} provides detailed analysis about the intra/inter-job interference between three applications on torus network. Section \ref{sec:discussion} talks about the idea of smart allocation strategy with job communication pattern awareness. Section \ref{sec:related_work} talks about the related work in the area and conclusion will be presented in Section \ref{sec:conclusion}.

@@

\begin{figure}[t!]
    \centering
    \begin{subfigure}[t]{0.22\textwidth}
        \centering
        \includegraphics[height=1.2in]{figs/appstudy/amg/amg_pip}
        \caption{Communication Topology}
        \label{fig: amg communication topology}
    \end{subfigure}%
    ~
    \begin{subfigure}[t]{0.22\textwidth}
        \centering
        \includegraphics[height=1.2in]{figs/appstudy/amg/amg_data_transfer}
        \caption{Data Volume}
        \label{fig: amg data trans}
    \end{subfigure}
    \caption{AMG. (a) rank-to-rank communication topology graph. (b) the data sent/received by each rank in AMG.}
\end{figure}

@@

\begin{figure}[t!]
    \centering
    \begin{subfigure}[t]{0.22\textwidth}
        \centering
        \includegraphics[height=1.2in]{figs/appstudy/cr/cr_pip}
        \caption{Communication Topology}
        \label{fig: cr communication topology}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{0.22\textwidth}
        \centering
        \includegraphics[height=1.2in]{figs/appstudy/cr/cr_data_transfer}
        \caption{Data Volume}
        \label{fig: cr data trans}
    \end{subfigure}
    \caption{CrystalRouter. (a) rank-to-rank communication topology graph. (b) the data sent/received by each rank in CrystalRouter. }
\end{figure}

@@

\begin{figure}[t!]
    \centering
    \begin{subfigure}[t]{0.22\textwidth}
        \centering
        \includegraphics[height=1.2in]{figs/appstudy/mg/mg_pip}
        \caption{Communication Topology}
        \label{fig: mg communication topology}
    \end{subfigure}
    ~
    \begin{subfigure}[t]{0.22\textwidth}
        \centering
        \includegraphics[height=1.2in]{figs/appstudy/mg/mg_data_transfer}
        \caption{Data Volume}
        \label{fig: mg data trans}
    \end{subfigure}
    \caption{MultiGrid. (a) rank-to-rank communication topology graph. (b) the data sent/received by each rank in MultiGrid.}
\end{figure}

@@

\subsection{Configuration Study }
\label{sec:config study}
